---
title: "Bloom spatiotemporal dynamics"
author: "Michelle Berry"
output: html_document
---

```{r global_options, echo = FALSE}

knitr::opts_chunk$set(
  fig.align = 'center', 
  warning = FALSE, 
  message = FALSE
)

```

```{r, echo=FALSE}
source("chabs-global.R")
```

# Figure 1
Goal is to create figure 1, showing pigment concentrations, toxin concentration, and 
proportion of cyanobacterial reads over time and stations. 

```{r figure1, fig.height=6, fig.width=6}

# Calculate relative abundance of Cyanobacteria at each date
cyano_abundance <- 
  erie_scale %>%
    tax_glom(taxrank = "Phylum") %>%                        # conglomerate OTUs to phylum level
    transform_sample_counts(function(x) {x/sum(x)} ) %>%    # transform to relative abundance
    subset_taxa(Phylum == "Cyanobacteria") %>%              # Subset to just Cyanobacteria
    psmelt() %>%                                            # melt phyloseq object
    rename(Cyanobacteria = Abundance) %>%
    select(Cyanobacteria, Date, Station)

# Merge cyanobacteria data with nutrient df
bloom_df <- 
  nutrient %>%
    left_join(cyano_abundance, by = c("Station", "Date")) %>%
    mutate(Phycocyanin = ifelse(Phycocyanin > 80, 80, Phycocyanin)) %>%   # lower extreme values to plot better
    select(Station, Date, Phycocyanin, Chla, ParMC, Cyanobacteria) %>%
    melt(id.vars = c("Station", "Date")) %>%
    order_dates_aug11()

# Make a faceted ggplot of the four bloom variables over time and grouped by station
bloom_plots <- ggplot(bloom_df, 
  aes(x = Date, y = value, group = Station, color = Station, shape = Station)
) +
  facet_grid(variable~., scales = "free_y") +
  geom_point(size = 1.3) +
  geom_line(size = 1) + 
  ylab("") +
  scale_x_discrete(
    breaks = c("7/8", "8/4", "9/2", "10/6"),
    labels = c("Jul", "Aug", "Sep", "Oct"),
    drop = FALSE
  ) +
  scale_color_manual(values = station_colors) + 
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 11),
    axis.title.x = element_blank()
  )


station_legend <- grab_legend(bloom_plots) 

bloom_plots

ggsave("../../Plots/Figure1.pdf", plot = bloom_plots, width = 6, height = 5) 
```

## Figure 1 Statistics
```{r fig.height=3.5, fig.width=10}
par(mfrow = c(1,3))
hist(nutrient$Chla)
hist(nutrient$Phycocyanin)
hist(nutrient$ParMC)
```

These distributions are very non-normal. Let's try log-scaling

```{r, fig.height=3.5, fig.width=10}
par(mfrow = c(1,3))
hist(log2(nutrient$Chla), xlab = "log Chla", main = "")
hist(log2(nutrient$Phycocyanin), xlab = "log Phycocyanin", main = "")
hist(log2(nutrient$ParMC), xlab = "log ParMC", main = "")
```

These look better, but parmc and phycocyanin both have zeroes so we would need to add a constant. Better to use a Spearman's test than a Pearson's test. 


```{r}
ggplot(nutrient, aes(x = Chla, y = Phycocyanin)) +
  geom_point() +
  scale_y_continuous(trans = log_trans()) +
  scale_x_continuous(trans = log_trans())

# Calculate spearman correlation between Chla and phycocyanin for all sites
cor.test(
  x = nutrient$Chla, 
  y = nutrient$Phycocyanin, 
  alternative = "two.sided", 
  method = "spearman"
)

```
It looks like there is a pretty close correlation between chl a and phycocyanin
measurements. 
    


```{r}

ggplot(nutrient, aes(x = Phycocyanin, y = ParMC)) +
  geom_point() +
  scale_y_continuous(trans = log_trans()) +
  scale_x_continuous(trans = log_trans())

cor.test(
  x = nutrient$Phycocyanin, 
  y = nutrient$ParMC, 
  alternative = "two.sided", 
  method = "spearman"
)

```
The statistical test shows that there are significant correlations between
pigments and toxin, but the plot shows that there are several dates in which 
the toxin levels are 0 but pigments are high, indicating that pigments are not
always predictive of toxicity. 


Do the nearshore and offshore sites have different bloom pigment concentrations?
```{r}
nutrient %>%
  group_by(Station) %>%
  summarise(median(Chla))

```


# Figure S1: Environmental variables
```{r fig.height= 10, fig.width = 10}


# Format
nutrient_df <- 
  sample_data(erie) %>%
    order_dates()

plot_nutrients <- function(df, nutrient, title, ylabs) {
  ggplot(df, aes_string(
    x = "Date",
    y = nutrient,
    group = "Station", 
    shape = "Station", 
    color = "Station")
  ) +
    scale_x_discrete(
      breaks = c("7/8", "8/4", "9/2", "10/6"),
      labels = c("Jul", "Aug", "Sep", "Oct"),
      drop = FALSE
    ) +
    geom_line(size = 0.8) +
    geom_point(size = 1.8) +
    ggtitle(title) + 
    xlab("") + 
    ylab(ylabs) +       
    scale_color_manual(values = station_colors) +
    theme(
        axis.title.y = element_text(size = 14),
        plot.title = element_text(size = 16, face = "bold")
    )
}

nutplot_vars <- c("SRP", "Nitrate", "Ammonia", "H2O2", "Temp", "Turbidity", "pH", "SpCond")
nutplot_titles <- c("SRP", "Nitrate", "Ammonia", "H2O2", "Temperature", "Turbidity", "pH", "SpCond")
nutplot_ylabs <- c(rep("ug/L", 3), "nM", "celsius", "NTU", "", "uS/cm")

nutplots <- list()

for (i in 1:length(nutplot_vars)) {
  nutplots[[i]] <- plot_nutrients(
    df = nutrient_df, 
    nutrient = nutplot_vars[i], 
    title = nutplot_titles[i],
    ylabs = nutplot_ylabs[i]
  )
}


legend <- grab_legend(nutplots[[1]])

nutplots_noleg <- lapply(nutplots, function(x){ x + theme(legend.position = "none")})
nutplots_noleg[[9]] <- legend

# multiplot
figs1 <- do.call("arrangeGrob", c(nutplots_noleg, ncol = 3))

ggdraw() +
  draw_plot(figs1) 

ggsave("../../Plots/FigureS1.pdf", width = 10, height = 10) 
```

