---
title: "Cyanobacterial community dynamics"
author: "Michelle Berry"
output: html_document
---

```{r global_options, echo = FALSE}

knitr::opts_chunk$set(
  fig.align = 'center', 
  warning = FALSE, 
  message = FALSE
)

```

```{r, echo=FALSE}
source("chabs-global.R")
```

# Figure 2: Cyano OTUs
In this figure we will display in part A a barplot with the relative abundance of
cyanobacteria genera over time and site. In part B we will break up these groups
into OTUs and show lineplots for each non-rare OTU over time (mean rel abundance > 0.0001) and site. 


```{r}
## Select only cyano OTUs that have mean relative abundace > 0.0005
n = 15000
thresh = 0.0005

# Calculate mean relative abundance for each OTU
tax_mean <- taxa_sums(erie_scale)/nsamples(erie_scale)

# Prune low abundance taxa using thresh as mean relative abundance
erie_prune_0001 <- 
  erie_scale %>%
    prune_taxa(tax_mean > thresh*n, .)


# Create a melted data frame of selected cyanobacteria OTUs
cyano_otus <- 
  erie_prune_0001 %>%
    transform_sample_counts(function(x) {x/sum(x)}) %>%
    subset_taxa(Class == "Cyanobacteria") %>%
    psmelt() %>%
    order_dates()
```

```{r include = FALSE}
na_dates <- data.frame(OTU = rep(levels(cyano_otus$Species), 3), Station = c(rep("nearshore1", 9), rep("nearshore2", 9), rep("offshore", 9)), Date = rep("8/11", 27))

cyano_otus1 <- 
  cyano_otus %>%
    full_join(na_dates, by = c("Date" = "Date", "Station" = "Station", "OTU" = "OTU")) %>%
    order_dates()
```

```{r}
################# Plot A #######################

cyano_genus <-
  cyano_otus %>%
    group_by(Genus, Date, Station) %>%
    summarize(Abundance = sum(Abundance)) %>%
    arrange(Genus) %>%
    order_dates()


plot2a <- ggplot(cyano_genus, aes(x = Date, y = Abundance, fill = Genus)) +
  facet_grid(~Station) +
  geom_bar(stat = "identity") +
  ylab("rel. abundance") + 
  scale_x_discrete(
      breaks = c("7/8", "8/4", "9/2", "10/6"),
      labels = c("Jul", "Aug", "Sep", "Oct"),
      drop = FALSE
    ) +
  scale_fill_manual(values = c("#377eb8", "#4daf4a", "#e41a1c", "#984ea3","#ff7f00")) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 9),
    strip.text = element_text(size = 11, face = "bold", hjust = 0),
    strip.background = element_blank()
  ) 

genus_legend <- grab_legend(plot2a)

plot2a <- plot2a + theme(legend.position = "none")
```



```{r}


##################### Plot B #############################

cyano_otu_names <- as.list(levels(cyano_otus$Species))
names(cyano_otu_names) <- levels(cyano_otus$Species)


# Generate a lineplot for each cyanobacteria with mean relative abundance > 0.0001
cyano_otu_plots <- lapply(cyano_otu_names, 
  function(otu) {
    df_otu <- filter(cyano_otus, OTU == otu)
    plot <- plot_otus(df = df_otu, otu = otu, taxrank = "Genus") 
    return(plot)
  }
)

# Generate legend
station_legend <- grab_legend(cyano_otu_plots$Otu00007)

cyano_otu_plots <- lapply(cyano_otu_plots, function(x) x + theme(legend.position = "none") )

plot2b <- arrangeGrob(
  cyano_otu_plots$Otu00007, cyano_otu_plots$Otu00177, cyano_otu_plots$Otu00005,
  cyano_otu_plots$Otu00044, cyano_otu_plots$Otu00304, cyano_otu_plots$Otu00037, 
  cyano_otu_plots$Otu00147, cyano_otu_plots$Otu00193, cyano_otu_plots$Otu00049,
  ncol = 3, nrow = 3
)


```


```{r fig.width=10, fig.height=8}
##################### Compile plots #####################################

ggdraw() +
  draw_plot(plot2a, x = 0.03, y = 0.7, width = .8, height = 0.28) +
  draw_plot(genus_legend, x = 0.83, y = .76, width = .16, height = .18) + 
  draw_plot(plot2b, x = 0.03, y = 0.02, width = .8, height = 0.64) +
  draw_plot(station_legend, x = 0.80, y = .5, width = .18, height = .18) + 
  draw_plot_label(c("A", "B"), c(0.02, 0.02), c(0.99, 0.68), size = 14) 

ggsave("../../Plots/Figure2.pdf", width = 10, height = 8) 

```

What is the most abundant genus?
```{r}
cyano_genus %>%
  group_by(Genus) %>%
  summarise(median(Abundance))

```

What if we look at each site separately?
```{r}
cyano_genus %>%
  group_by(Genus, Station) %>%
  summarise(median(Abundance))
```

Which Cyanobacteria are associated with each other? Spearman's rho
correlations are shown and significant
correlations are bolded. 
```{r}

# Create a phyloseq object of only cyanobacteria OTUs
cyano_physeq <- 
  erie_prune_0001 %>%
    transform_sample_counts(function(x) {x/sum(x)}) %>%
    subset_taxa(Class == "Cyanobacteria")

taxa_names(cyano_physeq) <- paste(
  tax_table(cyano_physeq)[ , "Genus"], 
  taxa_names(cyano_physeq)
)

# Run pairwise spearman correlation tests between all cyanobacteria genera.
cyano_corrs_spearman <- corr.test(
  t(otu_table(cyano_physeq)), 
  use = "complete", 
  method = "spearman", 
  adjust = "fdr"
)

emphasize.strong.cells(which(cyano_corrs_spearman$p < 0.05, arr.ind = TRUE))
pander(signif(cyano_corrs_spearman$r, digits = 2))

write.csv(cyano_corrs_spearman$p, "../../Tables/TableS1.csv", row.names = FALSE)

```
