---
title: "Nc-bacterial community composition analyses"
author: "Michelle Berry"
output: 
  html_document: 
    toc: yes
---

```{r global_options, echo = FALSE}

knitr::opts_chunk$set(
  fig.align = 'center', 
  warning = FALSE, 
  message = FALSE
)

```

```{r, echo=FALSE}
source("chabs-global.R")
library(pander)
```


# Figure 4


```{r}

# Subset to non-cyanobacteria and scale internally
ncbacteria <- 
  erie %>%
    subset_taxa(Class != "Cyanobacteria") %>%
    scale_reads(round = "round")


# Generate pcoa scores for each subset
ncbact_pcoa <- ordinate(
      physeq = ncbacteria,
      method = "PCoA",
      distance = "bray"
)

# Generate a df to plot pcoa for each subset
pcoa_df <- plot_ordination(
  physeq = ncbacteria,
  axes = 1:3,
  ordination = ncbact_pcoa,
  justDF = TRUE
)

pcoa_df <- pcoa_df %>% 
  rename(PC1 = Axis.1, PC2 = Axis.2, PC3 = Axis.3) %>%
  order_dates() %>%
  # Flip orientation of PC2 for Cyanobacteria (does not affect interpretation)
  mutate(PC2 = -PC2) 


# Generate relative, lingoes-corrected eigenvalues for PC1 and PC2
pcs <- c(
  PC1 = signif(ncbact_pcoa$values$Rel_corr_eig[1]*100, 3),
  PC2 = signif(ncbact_pcoa$values$Rel_corr_eig[2]*100, 3),
  PC3 = signif(ncbact_pcoa$values$Rel_corr_eig[3]*100, 3)
)

```

```{r}
# Function to create a plot of time (x-axis) vs PC scores (y-axis)
plot_pcts <- function(df, pc, eigs) {
  ggplot(df, 
    aes_string(x = "Date", y = pc, group = "Station", color = "Station", shape = "Station")) +
      geom_point(size = 2.5) +
      geom_line(size = 1.1) + 
      scale_color_manual(values = station_colors) +
      scale_x_discrete(
        breaks = c("7/8", "8/4", "9/2", "10/6"),
        labels = c("Jul", "Aug", "Sep", "Oct"),
        drop = FALSE
      ) +  
      ylab(paste(pc, " ", eigs[pc], "%")) +
      theme(
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        plot.title = element_text(face = "bold", size = 16),
        legend.position = "none",
        plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "cm")
      ) 
}

# Non-cyano PC time series plots
ncbact_pcs <- named_list("PC1", "PC2", "PC3")

ncbact_pc_plots <- lapply(ncbact_pcs,
  function(x) {
    plot_pcts(pcoa_df, x, eigs = pcs)
  }
)

```


```{r}

# Plot for pH
ph_plot <- ggplot(pcoa_df, aes(x = PC1, y = pH)) +
  geom_point(size = 2.5, aes(shape = Station, color = Station)) +
  scale_color_manual(values = station_colors) +
  geom_smooth(method = "lm", color = "black") +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10)
  ) 

legend <- grab_legend(ph_plot)
ph_plot <- ph_plot + theme(legend.position = "none")

# Plot for Temperature
temp_plot <- ggplot(pcoa_df, aes(x = PC2, y = Temp)) +
  geom_point(size = 2.5, aes(shape = Station, color = Station)) +
  scale_color_manual(values = station_colors) +
  geom_smooth(method = "lm", color = "black") +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "none"
  ) +
  ylab("Temperature")

# Plot for SpCond
spcond_plot <- ggplot(pcoa_df, aes(x = PC3, y = SpCond)) +
  geom_point(size = 2.5, aes(shape = Station, color = Station)) +
  scale_color_manual(values = station_colors) +
  geom_smooth(method = "lm", color = "black") +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "none"
  ) +
  ylab("Specific Conductivity")

```

## Compiled plot
```{r fig.height=7, fig.width=12}

ggdraw() +
  ## non-cyano
  draw_plot(ncbact_pc_plots$PC1, x = 0.02, y = 0.61, width = 0.3, height = 0.35) +
  draw_plot(ncbact_pc_plots$PC2, x = 0.34, y = 0.61, width = 0.3, height = 0.35) +
  draw_plot(ncbact_pc_plots$PC3, x = 0.66, y = 0.61, width = 0.3, height = 0.35) +
  draw_plot(ph_plot, x = 0.02, y = 0.22, width = 0.3, height = 0.35) +
  draw_plot(temp_plot, x = 0.34, y = 0.22, width = 0.3, height = 0.35) +
  draw_plot(spcond_plot, x = 0.66, y = 0.22, width = 0.3, height = 0.35) +
  draw_plot_label(c("A", "B", "C", "D", "E", "F"),  c(0.02, 0.02, 0.34, 0.34, 0.65, 0.65), c(0.97, 0.61, 0.97, 0.61, 0.97, 0.61), size = 14) +
  draw_plot(legend, x = 0.02, y = 0.02, width = 0.2, height = 0.2)

  
ggsave("../../Plots/Figure4.pdf", width = 12, height = 7) 
```

# Figure 4 Statistics

How many PCoA dimensions should we examine? 

```{r}
#pdf("../../Plots/FigureS4.pdf", width = 6, height = 4) 
plot(1:nrow(ncbact_pcoa$values), ncbact_pcoa$values$Rel_corr_eig, xlab = 
       "Principal Coordinate", ylab = "Relative eigenvalue")

#dev.off()
```

How good is my PCoA in three dimensions?
```{r}
pcoa_dist <- as.vector(dist(ncbact_pcoa$vectors[,1:3]))
bray_dist <- as.vector(phyloseq::distance(ncbacteria, method = "bray"))

plot(bray_dist, pcoa_dist) +
 abline(lm(pcoa_dist~bray_dist), col = "blue")

summary(lm(pcoa_dist~bray_dist))
```


```{r}
# Histogram of bray-curtis distance
hist(bray_dist)
```

What is the max bray-curtis between time points for each site? 
```{r}

# Nearshore 1 
n1 <- subset_samples(erie_scale, Station == "nearshore1")
n1_bdist <- phyloseq::distance(n1, method = "bray")
max(as.matrix(n1_bdist)[1,])


# Nearshore 2
n2 <- subset_samples(erie_scale, Station == "nearshore2")
n2_bdist <- phyloseq::distance(n2, method = "bray")
max(as.matrix(n2_bdist)[1,])

# Offshore 
o <- subset_samples(erie_scale, Station == "offshore")
o_bdist <- phyloseq::distance(o, method = "bray")
max(as.matrix(o_bdist)[1,])


```

What is the bray-curtis between the first and last time points
```{r}

# Nearshore 1 
n1 <- subset_samples(erie_scale, Station == "nearshore1" & Date %in% c("6/16", "10/27"))
n1_bdist <- phyloseq::distance(n1, method = "bray")
n1_bdist

# Nearshore 2
n2 <- subset_samples(erie_scale, Station == "nearshore2" & Date %in% c("6/16", "10/27"))
n2_bdist <- phyloseq::distance(n2, method = "bray")
n2_bdist

# Offshore 
o <- subset_samples(erie_scale, Station == "offshore" & Date %in% c("6/16", "10/27"))
o_bdist <- phyloseq::distance(o, method = "bray")
o_bdist

```



## Linear models
```{r, fig.height=10, fig.width=10}

# Variables to include in cyano models
cyano_vars <- c("Nitrate", "SRP", "Temp", "H2O2", "SpCond", "Ammonia", "Turbidity", "TP", "Days")


# Examine distributions of potential env variables to normalize
par(mfrow = c(4,3))
for (var in cyano_vars) {
  hist(nutrient[ ,var], main = "", xlab = var, ylab = "" )
}
```

```{r}
# Adjusted variables to include in cyano models
cyano_vars <- c("sqrt(Nitrate)", "log(SRP)", "Temp", "log(H2O2)", "SpCond", "log(Ammonia)", "log(Turbidity)", "log(TP)")


# Variables to include in nc-bacteria models
non_cyano_vars <- c(cyano_vars, "pH",  "logParMC", "logChla", "logPhyco")


# Impute SpCond values for nearshore 1 on Sep 2 and Sep 8 with value for nearshore 2

# Change 9/2 value
pcoa_df$SpCond[pcoa_df$Date == "9/2" & pcoa_df$Station == "nearshore1"] <- 
  pcoa_df$SpCond[pcoa_df$Date == "9/2" & pcoa_df$Station == "nearshore2"]
# Change 9/8 value
pcoa_df$SpCond[pcoa_df$Date == "9/8" & pcoa_df$Station == "nearshore1"] <- 
  pcoa_df$SpCond[pcoa_df$Date == "9/8" & pcoa_df$Station == "nearshore2"]


# Add log scaled variables
pcoa_df <- pcoa_df %>%
  mutate(logParMC = log(ParMC + 0.1)) %>%
  mutate(logChla = log(Chla)) %>%
  mutate(logPhyco = log(Phycocyanin + 0.1))
  
```


```{r env variables}

get_detrend_models <- function(response, dat) {
  fit <- lm(reformulate(termlabels = "Days", response = response), data = dat)
  return(fit$residuals)
}

get_time_models <- function(vars, response, dat) {
  models <- list()
  for (var in vars) {
    models[[var]] <- lm(reformulate(termlabels = c(var, "Days"), response = response), data = dat)
  }
  return(models)
}

get_notime_models <- function(vars, response, dat) {
  models <- list()
  for (var in vars) {
    models[[var]] <- lm(reformulate(termlabels = var, response = response), data = dat)
  }
  return(models)
}
```

```{r}
ncbact_pcs <- named_list("PC1", "PC2", "PC3")

pcoa_df$PC1detrend <- get_detrend_models(response = "PC1", dat = pcoa_df)
pcoa_df$PC2detrend <- get_detrend_models(response = "PC2", dat = pcoa_df)
pcoa_df$PC3detrend <- get_detrend_models(response = "PC3", dat = pcoa_df)

# Get the simple linear models for each variable along each PC
non_cyano_models <- lapply(ncbact_pcs, function(x) {
  if (x == "PC2") {
    get_notime_models(non_cyano_vars, x, dat = pcoa_df)
  } else {
    get_time_models(non_cyano_vars, x, dat = pcoa_df)
  }
   
})

# Get rsquared for each model
rsquared <- lapply(non_cyano_models, function(x) {
  lapply(x, function(y) {
    r2 <- summary(y)$r.squared
    return(r2)
  })
})

rsquared_df <- rbind(data.frame(rsquared$PC1), data.frame(rsquared$PC2), data.frame(rsquared$PC3))
rsquared_df$PC <- c("PC1", "PC2", "PC3")

pander(t(as.matrix(rsquared_df)))


# Get rsquared for each model
pvalue <- lapply(non_cyano_models, function(x) {
  lapply(x, function(y) {
    p <- summary(y)$coefficients[2,4]
    return(p)
  })
})

pvalue_df <- rbind(data.frame(pvalue$PC1), data.frame(pvalue$PC2), data.frame(pvalue$PC3))
pvalue_df$PC <- c("PC1", "PC2", "PC3")

pander(t(as.matrix(pvalue_df)))
```

### Best models
```{r}
pc1_ph <- lm(PC1 ~ pH + Days, data = pcoa_df)
summary(pc1_ph)
pc1_chla <- lm(PC1 ~ logChla + Days, data = pcoa_df)
summary(pc1_chla)
pc2 <- lm(PC2 ~ Temp + Turbidity + SpCond + SRP, data = pcoa_df)
summary(pc2)
pc3 <- lm(PC3 ~ SpCond, data = pcoa_df)
summary(pc3)
```


### Distribution of model residuals
```{r fig.height=4, fig.width=10}
#pdf("../../Plots/FigureS5.pdf", width = 10, height = 3) 
par(mfrow = c(1, 4))
hist(residuals(pc1_ph))
hist(residuals(pc1_chla))
hist(residuals(pc2))
hist(residuals(pc3))

#dev.off()
```
All distributions look normal enough.



### Temporal autocorrelation of model residuals:
```{r, fig.width=14, fig.height=14}

#pdf("../../Plots/FigureS6.pdf", width = 14, height = 14) 

par(mfrow = c(4, 3))
# Look at temporal autocorrelation
n1 <- which(pcoa_df$Station == "nearshore1")
n2 <- which(pcoa_df$Station == "nearshore2")
o <- which(pcoa_df$Station == "offshore")

n1_ph <- which(!is.na(pcoa_df$pH) & pcoa_df$Station == "nearshore1")
n2_ph <- which(!is.na(pcoa_df$pH) & pcoa_df$Station == "nearshore2")
o_ph <- which(!is.na(pcoa_df$pH) & pcoa_df$Station == "offshore")


acf(residuals(pc1_ph)[n1_ph], main = "PC1 ~ pH; nearshore 1")
acf(residuals(pc1_ph)[n2_ph], main = "PC1 ~ pH; nearshore 2")
acf(residuals(pc1_ph)[o_ph], main = "PC1 ~ pH; offshore")


acf(residuals(pc1_chla)[n1], main = "PC1 ~ chla; nearshore 1")
acf(residuals(pc1_chla)[n2], main = "PC1 ~ chla; nearshore 2")
acf(residuals(pc1_chla)[o], main = "PC1 ~ chla; offshore")


acf(residuals(pc2)[n1], main = "PC2 ~ Temp; nearshore 1")
acf(residuals(pc2)[n2], main = "PC2 ~ Temp; nearshore 2")
acf(residuals(pc2)[o], main = "PC2 ~ Temp; offshore")


acf(residuals(pc3)[n1], main = "PC3 ~ SpCond; nearshore 1")
acf(residuals(pc3)[n2], main = "PC3 ~ SpCond; nearshore 2")
acf(residuals(pc3)[o], main = "PC3 ~ SpCond; offshore")

#dev.off()

```

Partial autocorrelation function:

```{r fig.width=14, fig.height=14}
par(mfrow = c(4, 3))


pacf(residuals(pc1_ph)[n1_ph], main = "PC1 ~ pH; nearshore 1")
pacf(residuals(pc1_ph)[n2_ph], main = "PC1 ~ pH; nearshore 2")
pacf(residuals(pc1_ph)[o_ph], main = "PC1 ~ pH; offshore")

pacf(residuals(pc1_chla)[n1], main = "PC1 ~ pH; nearshore 1")
pacf(residuals(pc1_chla)[n2], main = "PC1 ~ pH; nearshore 2")
pacf(residuals(pc1_chla)[o], main = "PC1 ~ pH; offshore")

pacf(residuals(pc2)[n1], main = "PC2 ~ Temp; nearshore 1")
pacf(residuals(pc2)[n2], main = "PC2 ~ Temp; nearshore 2")
pacf(residuals(pc2)[o], main = "PC2 ~ Temp; offshore")


pacf(residuals(pc3)[n1], main = "PC3 ~ SpCond; nearshore 1")
pacf(residuals(pc3)[n2], main = "PC3 ~ SpCond; nearshore 2")
pacf(residuals(pc3)[o], main = "PC3 ~ SpCond; offshore")


```




### Cross validation error
leaving out one time point at a time:
```{r}
dates <- levels(pcoa_df$Date)
loocv <- rep(0, length(dates))
names(loocv) <- dates

# PC1
for (date in dates) {
  train <- filter(pcoa_df, Date != date)
  test <- filter(pcoa_df, Date == date)
  fit <- lm(PC1 ~ pH + Days, data = train)
  pred <- predict(fit, test)
  loocv[date] <- mean((test$PC1 - pred)^2)
}
mean(loocv[-18])


# PC1
for (date in dates) {
  train <- filter(pcoa_df, Date != date)
  test <- filter(pcoa_df, Date == date)
  fit <- lm(PC1 ~ logChla + Days, data = train)
  pred <- predict(fit, test)
  loocv[date] <- mean((test$PC1 - pred)^2)
}
mean(loocv)

# PC2
for (date in dates) {
  train <- filter(pcoa_df, Date != date)
  test <- filter(pcoa_df, Date == date)
  fit <- lm(PC2 ~ Temp, data = train)
  pred <- predict(fit, test)
  loocv[date] <- mean((test$PC2 - pred)^2)
}
mean(loocv)

# PC3
for (date in pcoa_df$Date) {
  train <- filter(pcoa_df, Date != date)
  test <- filter(pcoa_df, Date == date)
  fit <- lm(PC3 ~ SpCond, data = train)
  pred <- predict(fit, test)
  loocv[date] <- mean((test$PC3 - pred)^2)
}
mean(loocv)
```

